Bayesian MCMC diagnostics and summarisation using Tracer 1.6

### 2017-07-27 manuscript/tracer16.tex
add example 2

Tracer also provides demographic reconstruction resulting in a graphical plot to estimate the distribution of effective population sizes over time, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, and logistic growth (?). %citation?
Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
``Lineages Through Time'' (?) plots quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
All these models are available in BEAST \citep{drummond2007beast}.

@@

Figure~\ref{fig:4tabs} demonstrates a snapshot of posterior summaries in Tracer for a BEAST \citep{drummond2007beast} run created from a 898 bp alignment of sequences of 12 species of primates (http://beast.bio.ed.ac.uk/tutorials?). %data citation
%using HKY + $\Gamma$ model, uncorrelated lognormal relaxed molecular clock, and Yule prior on the tree.
In this analysis, we furthermore specify three calibrations on the phylogenetic tree to estimate time to the most recent common ancestor (TMRCA) based on our prior fossil knowledge.

The first TMRCA prior, called ``ingroup'', includes all species except the \textit{Lemur} which will form the outgroup in the tree. %monophyletic
The second group is ``Human-Chimp'' that contains only \textit{Homo sapiens} and \textit{Pan}. 
The last consists of everything except \textit{Lemur}, \textit{Saimiri} and \textit{Tarsius} to create the hominoid-cercopithecoid split, which is named as ``HomiCerco''. 
Since we know the most recent common ancestor of humans and chimps may live on about $5-7$ million years ago, a normal distribution centered at 6 million years with a standard deviation of 0.5 million years is assigned to the ``Human-Chimp'' prior. 
Another normal distribution of $24 \pm 0.5$ (million years) is likewise defined for ``HomiCerco''.

@@

Selecting the ``treeModel.rootHeight'' parameter gives the marginal posterior distribution of the age of the root of entire tree. In the example in Figure~\ref{fig:4tabs}.a, we can conclude the root of the tree is most likely between about 39.01 and 63.67 million years old with the mean of 50.48 million years.

According to the kernel density estimation of the two TMRCA parameters (``Human-Chimp'' in blue and ``HomiCerco'' in yellow) in Figure~\ref{fig:4tabs}.b, these two parameters that we used to calibrate the tree explicitly have posterior distributions very similar to the prior distributions that we specified.

The marginal posterior densities of ``treeModel.rootHeight'' and ``ingroup'' are nearly overlapped each other. If we switch to Figure~\ref{fig:4tabs}.c, their joint-marginal distribution plot is almost lined up together, which indicates ??. But this isn't an issue here, because the ``ingroup'' only excludes one outgroup species from everyone else. %TODO

@@

\subsection*{Demographic reconstruction}

The second analysis uses an 1698bp alignment of 21 sequences from the Influenza A virus (?). %data citation
The piecewise-constant Bayesian skyline is selected for the tree prior, where the number of groups is reduced to 5 to avoid over-parameterize the model.

\begin{figure}[ht]
%\includegraphics[width=.5\textwidth]{./figures/fluBSP.pdf}  
\subfigure[Bayesian Skyline]{\includegraphics[width = .23\textwidth]{./figures/fluBSP.pdf}}
\subfigure[Lineages Through Time]{\includegraphics[width = .23\textwidth]{./figures/fluLTT.pdf}}
\caption{The plot (a) performs Bayesian Skyline reconstruction to estimate population sizes over time and (b) Lineages Through Time , from ``Flu'' data, where time starts from year 1997 at the 0 position and the unit is year.}
\label{fig:flu}
\end{figure}

>>>>

Tracer also provides demographic reconstruction resulting in a graphical plot to estimate the distribution of effective population sizes over time, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, and logistic growth (?). %citation?
Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian coalescent skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
``Lineages Through Time'' (?) plots quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
All these models are available in BEAST \citep{drummond2007beast}.

@@

Figure~\ref{fig:4tabs} demonstrates a snapshot of posterior summaries in Tracer for a BEAST \citep{drummond2007beast} run created from a 898 bp alignment of sequences of 12 species of primates (http://beast.bio.ed.ac.uk/tutorials?). %data citation
%using HKY + $\Gamma_4$ model, uncorrelated lognormal relaxed molecular clock, and Yule prior on the tree.
In this analysis, we furthermore specify three calibrations on the phylogenetic tree to estimate time to the most recent common ancestor (TMRCA) based on our prior fossil knowledge.

The first TMRCA prior, called ``ingroup'', includes all species except the \textit{Lemur} which will form the outgroup in the tree. %monophyletic
The second group is ``Human-Chimp'' that contains only \textit{Homo sapiens} and \textit{Pan}. 
The last consists of everything except \textit{Lemur}, \textit{Saimiri} and \textit{Tarsius} to create the hominoid-cercopithecoid split, which is named as ``HomiCerco''. 
Since we know the most recent common ancestor of humans and chimps may live about $5-7$ million years ago, a normal distribution centered at 6 million years with a standard deviation of 0.5 million years is assigned to the ``Human-Chimp'' prior. 
Another normal distribution of $24 \pm 0.5$ (million years) is likewise defined for ``HomiCerco''.

@@

Selecting the ``treeModel.rootHeight'' parameter gives the marginal posterior distribution of the age of the root of entire tree in the example in Figure~\ref{fig:4tabs}.a. We can conclude the root of the tree is most likely between about 39.01 and 63.67 million years old expressed by the 95\% HPD interval with the mean of 50.48 million years.

Figure~\ref{fig:4tabs}.b shows the kernel density estimation of the two TMRCA parameters that ``Human-Chimp'' is blue and ``HomiCerco'' is yellow. These two parameters that we used to calibrate the tree explicitly have posterior distributions very similar to the prior distributions that we specified.

The marginal posterior densities of ``treeModel.rootHeight'' and ``ingroup'' are nearly overlapped each other, when we switch to Figure~\ref{fig:4tabs}.c. Their joint-marginal distribution plot is almost lined up together, which indicates ??. But this isn't an issue here, because the ``ingroup'' only excludes one outgroup species from everyone else. %TODO how to explain this problem further?

@@

\subsection*{Demographic reconstruction and temporal analysis}

In the second example, there are two analyses conducted using an 1698bp alignment of 21 sequences from the Influenza A virus (H5N1?) collected between 1997 and 2005 (?). %data citation
To perform Bayesian coalescent skyline plot in Figure~\ref{fig:flu}.a, we exploit Tracer to exact past population dynamics from both MCMC log and tree log produced by a BEAST \citep{drummond2007beast} run.
%using piecewise-constant Bayesian skyline tree prior with 5 groups in order to avoid over-parameterize the model.
The plot reveals the growth of the effective population sizes of Influenza infections during 8 years, which obviously involves an epidemic break-out period between 2002 and 2004.

\begin{figure}[ht]
%\includegraphics[width=.5\textwidth]{./figures/fluBSP.pdf}  
\subfigure[Bayesian Coalescent Skyline]{\includegraphics[width = .23\textwidth]{./figures/fluBSP.pdf}}
\subfigure[Lineages Through Time]{\includegraphics[width = .23\textwidth]{./figures/fluLTT.pdf}}
\caption{(a) Bayesian Coalescent Skyline plot and (b) Lineages Through Time plot for the Influenza A virus data.}
\label{fig:flu}
\end{figure}

Lineages Through Time in Figure~\ref{fig:flu}.b, similar curve

diversification has been gradually developed before 2003's break-out, but it also looks stopped after 2004.
